Speed and strength of epidemic intervention

### 2019-11-13 doc/speedstrength.tex
trying to work

An epidemic can be characterized by its \emph{speed} and \emph{strength}.
The speed of an epidemic, often characterized by the exponential growth rate $r$, measures how \emph{fast} an epidemic grows, at the population level. 
The strength of an epidemic, often characterized by the reproductive number $\RR$, measures how many new cases are caused by a typical \emph{individual} case.
Knowing the speed and strength of an epidemic allows predictions about the course of the epidemic and the effectiveness of intervention strategies.

Much research has prioritized estimates of the reproductive number, because it has a threshold value, $\RR=1$ that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts. 
The insight that a case must on average cause at least one new case under good conditions for a disease persist goes back $>100$ years \citep{ross1911prevention};
the idea of averaging by defining a `typical' case was formalized 30 years ago \citep{diekmann1990definition}.
$\RR$ is also of interest because it provides a \emph{prima facie} prediction about the total \emph{size} of an epidemic \citep{anderson1991infectious, ma2006generality, miller2012note}.

Like epidemic strength, epidemic speed measures how disease spreads in the population. 
Here, we generalize the idea of a threshold for successful intervention by measuring an intervention's ``strength'' on the same scale as the reproductive number. 
We then show that we can likewise measure an intervention's ``speed'', and that there is a duality between the threshold $\RR=1$ and a corresponding minimal intervention strength, and the threshold $r=0$ and a corresponding minimal intervention speed. 
We argue that the historical primacy of $\RR$ over $r$ is partly artificial, and discuss cases where strength provides the better framing for practical disease questions and cases where speed does.

>>>>

An epidemic can be described by its \emph{speed} and \emph{strength}.
The speed of an epidemic, often characterized by the exponential growth rate $r$, measures how \emph{fast} an epidemic grows, at the population level. 
The strength of an epidemic, often characterized by the basic reproductive number $\RR_0$, measures how many new cases are caused by a typical \emph{individual} case in a fully susceptible population \citep{anderson1991infectious, diekmann1990definition}.
Knowing the speed and strength of an epidemic allows predictions about the course of the epidemic and the effectiveness of intervention strategies.

Much research has prioritized estimates of the reproductive number, because it has a threshold value, $\RR_0=1$ that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts.
The insight that a case must on average cause at least one new case under good conditions for a disease persist goes back $>100$ years \citep{ross1911prevention};
the idea of averaging by defining a `typical' case was formalized 30 years ago \citep{diekmann1990definition}.
$\RR_0$ is also of interest because it provides a \emph{prima facie} prediction about the total \emph{size} of an epidemic \citep{anderson1991infectious, ma2006generality, miller2012note}.

The study of measles dynamics provides a key example for which the characterization of an epidemic via the reproduction number has been successful.
One of the earliest estimates of the measles $\RR_0$ comes from \cite{anderson1982directly} nearly 40 years ago, who gave a range of $14 - 18$.
This, combined with the mean age of infection (4.6 years) and the mean age of vaccination (2.2 years), allowed them to predict that approximately 96\% of the new birth cohort had to be vaccinated to eradicate measles.

Like epidemic strength, epidemic speed measures how disease spreads in the population. 
\swp{Ebola}

\swp{Eaton and Hallett}

Here, we generalize the idea of a threshold for successful intervention by measuring an intervention's ``strength'' on the same scale as the reproductive number. 
We then show that we can likewise measure an intervention's ``speed'', and that there is a duality between the threshold $\RR=1$ and a corresponding minimal intervention strength, and the threshold $r=0$ and a corresponding minimal intervention speed. 
We argue that the historical primacy of $\RR$ over $r$ is partly artificial, and discuss cases where strength provides the better framing for practical disease questions and cases where speed does.
